Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-04-30 paper.tex
Up interacciones WExTime AL

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\section{Theoretical views on market justice, inequality perception and
meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}

@@

Market justice has been discussed in the literature as a normative
principle that legitimates the distribution of economic rewards based on
individual merit and effort. The most detailed theoretical account of
this concept is provided by Lane
(\citeproc{ref-lane_market_1986}{1986}), who defines market justice as a
system of ``earned deserts,'' whereby individuals are seen as entitled
to their outcomes insofar as these are a product of personal
productivity and competition. This view directly contrasts with
political justice, which emphasizes redistributive mechanisms based on
the principles of equality and need, and is often materialized through
welfare policies. For Lane (\citeproc{ref-lane_market_1986}{1986}),
market justice relies on the conception that markets are neutral and
self-regulating arenas, where individuals are treated fairly because
they face the same formal rules of engagement. What matters is not
whether everyone ends up with the same, but that everyone has the chance
to compete, and that success reflects effort and ability. Consequently,
the legitimacy of market justice stems from the assumption that
inequalities are not only inevitable but fair---so long as the rules are
transparent and opportunities are open. These beliefs align with a
broader ideological framework that values individual autonomy, private
property, and limited government intervention. In this way, market
justice provides a moral justification for inequality by framing it as a
necessary outcome of personal responsibility and societal efficiency.

Empirical studies have developed different strategies to capture and
operationalize market justice preferences. A common approach in the
literature is to gauge attitudes toward the legitimacy of inequality in
specific domains, especially when linked to income differences. Kluegel
and Smith (\citeproc{ref-kluegel_beliefs_1981}{1981}) and later Kluegel
et al. (\citeproc{ref-kluegel_legitimation_1999}{1999}) laid the
groundwork for measuring beliefs in market justice through survey items
that assess whether people support economic inequality when it reflects
personal effort and skills. Over time, this approach has been extended
beyond income to include other market-mediated outcomes, such as
education, healthcare, and/or pensions. For example, Von Dem Knesebeck
et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
citizens consider it fair that individuals with higher incomes can
access better healthcare, while Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
context of education in Australia. These studies usually rely on a
survey item asking respondents to evaluate the fairness of income-based
access to welfare services, allowing for comparing justice perceptions
across different contexts. Some research, like that of Lindh
(\citeproc{ref-lindh_public_2015}{2015}) and Svallfors
(\citeproc{ref-svallfors_political_2007}{2007}), uses these indicators
in combination to measure broader preferences for market-based
distribution in welfare systems. More recently, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. These different operationalizations all aim to
capture the extent to which individuals accept inequality when it is
framed as a reflection of market outcomes.

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is strong empirical evidence suggesting
that individuals in more advantaged socioeconomic positions---those with
higher income, education, and occupational status---are more likely to
endorse market-based distribution as fair and legitimate
(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). This tendency
reflects not only material self-interest but also a broader moral
economy, in which winners of the market system internalize
justifications for the status quo. At the same time, normative and
ideological orientations---such as conservative, liberal, or neoliberal
values---are consistently associated with greater support for
meritocratic principles and skepticism toward redistribution. This is
particularly visible in contexts such as Chile and Australia, where
studies by Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) and Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) show that those with
right-leaning political preferences tend to view income-based access to
welfare services as legitimate. Beyond individual-level characteristics,
country-level institutions also play a central role. In liberal welfare
regimes like those of the United States or the United Kingdom, market
justice preferences are more widespread, while in coordinated or
social-democratic regimes---such as Sweden or Germany---citizens are
generally more critical of market-based inequalities
(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
that market justice is not a fixed belief but a contextually shaped
orientation, mediated by structural conditions, cultural values, and the
historical development of welfare institutions.

@@

Perceptions of inequality have been associated to attitudes about market
justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al., 1995};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research indicates that
lower perceived inequality can reinforce support for market-based
distributive arrangements by suggesting that the system is fair and that
outcomes reflect effort and ability (\citeproc{ref-kuhn_eye_2011}{Kuhn,
2011}). In contrast, when inequality is perceived as excessive or
structurally determined, individuals could question the legitimacy of
market justice and become more supportive of redistributive policies
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}).

Perceptions of inequality refer to individuals' subjective evaluations
of the extent, causes, and consequences of income and wealth
disparities. Unlike objective measures such as the Gini index, perceived
inequality captures how individuals make sense of distributive
hierarchies in their everyday lives, shaped by reference groups, social
comparisons, and information environments
(\citeproc{ref-garcia-castro_perceiving_2020}{García-Castro et al.,
2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016}).
Scholars have proposed multiple dimensions of perceived inequality,
including its magnitude (how significant are the gaps), vertical
structure (between which groups), trend over time (increasing or
decreasing), and legitimacy (whether it is just or not)
(\citeproc{ref-engelhardt_what_2018}{Engelhardt \& Wagener, 2018};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}). These dimensions encompass both cognitive and normative aspects
of perceptions of inequality and can vary across societies and social
groups, depending on exposure, ideology, and personal experience
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
2018}).

@@

Meritocracy constitutes a central ideological framework for legitimizing
different types od social inequality, for instance through market
justice beliefs. Rooted in the belief that rewards and positions should
be allocated based on individual effort and talent, meritocracy operates
as a normative ideal and a descriptive belief about how society
functions. As initially conceptualized by Michael Young
(\citeproc{ref-young_rise_1962}{1962}), the term was meant to critique a
system in which merit-based stratification becomes a new form of
inequality. However, over time, meritocracy has been widely supported in
many societies as a fair and desirable principle of distribution,
particularly within liberal democracies and market-oriented economies
(\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
perspective, the belief in meritocracy is more than a cognitive
assessment; it reflects a moral lens through which individuals interpret
inequality. People who believe that success results from hard work and
talent are more likely to view social and economic disparities as
legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
Conversely, if they see outcomes as driven by luck, social origin, or
systemic barriers, inequality is more likely to be perceived as unjust.
This distinction becomes crucial in societies with persistent structural
inequality, where public narratives often emphasize personal
responsibility and merit while overlooking entrenched disadvantages.

@@

These dimensions of meritocracy reflect how respondents perceive
society's distributive logic, regardless of whether they personally
endorse meritocratic principles. This distinction aligns with recent
findings indicating that individuals distinguish between how merit is
perceived in society and how it should ideally operate, which in turn
shapes their preferences for redistribution and justice
(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
2025}) Meritocratic beliefs serve as symbolic justifications for unequal
outcomes, particularly when access is stratified by income or social
opportunity. Prior studies have shown that individuals who perceive
higher levels of meritocracy tend to express stronger support for
unequal distributions that reflect market outcomes
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).

In addition to influencing individual attitudes toward inequality,
meritocratic beliefs can contribute to social division and the
stigmatization of disadvantaged groups. Recent research has demonstrated
that exposure to meritocratic narratives can reinforce the belief that
poverty results from individual failings rather than systemic
conditions, reducing support for redistributive measures and increasing
the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
et al., 2023}). This reinforces negative stereotypes and reduces empathy
toward individuals from lower socioeconomic backgrounds. Moreover,
Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argue
that meritocratic narratives can serve as feedback mechanisms that shape
public opinion and well-being by framing individuals' understanding of
welfare outcomes as deserved or undeserved within existing institutional
structures. This psychological mechanism highlights the normative power
of meritocracy in stabilizing unequal systems by shaping political
attitudes and personal perceptions of success and failure.

@@

\(H1\): Greater perceived economic inequality is associated with
stronger market justice preferences.

\(H2\): Stronger meritocratic beliefs are associated with stronger
market justice preferences.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3367}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3367}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2143}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1122}}@{}}
\caption{Dependent variables for the first wave
(2016)}\label{tbl-summary1}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Label
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Stats / Values
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Freqs (\% of Valid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Valid
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Label
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Stats / Values
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Freqs (\% of Valid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Valid
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Health distributive justice &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
554 (37.5\%)\\
713 (48.3\%)\\
61 ( 4.1\%)\\
131 ( 8.9\%)\\
18 ( 1.2\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1477\\
(100.0\%)\strut
\end{minipage} \\
Pension distributive justice &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
425 (28.8\%)\\
704 (47.7\%)\\
106 ( 7.2\%)\\
219 (14.8\%)\\
23 ( 1.6\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1477\\
(100.0\%)\strut
\end{minipage} \\
Education distributive justice &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
519 (35.1\%)\\
764 (51.7\%)\\
71 ( 4.8\%)\\
112 ( 7.6\%)\\
11 ( 0.7\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1477\\
(100.0\%)\strut
\end{minipage} \\
Market justice preferences & \begin{minipage}[t]{\linewidth}\raggedright
Mean (sd) : 2 (0.8)\\
min \textless{} med \textless{} max:\\
1 \textless{} 2 \textless{} 5\\
IQR (CV) : 1 (0.4)\strut
\end{minipage} & 12 distinct values &
\begin{minipage}[t]{\linewidth}\raggedright
1477\\
(100.0\%)\strut
\end{minipage} \\
\end{longtable}

@@

Intercept                            & $1.968^{***}$  & $1.973^{***}$  & $1.985^{***}$  & $1.982^{***}$  & $2.122^{***}$  & $1.408^{***}$  & $1.499^{***}$  \\
                                     & $(0.021)$      & $(0.032)$      & $(0.032)$      & $(0.032)$      & $(0.069)$      & $(0.096)$      & $(0.114)$      \\

@@

Wave                                 &                & $-0.075^{***}$ & $-0.078^{***}$ & $-0.079^{***}$ & $-0.079^{***}$ & $-0.079^{***}$ & $-0.079^{***}$ \\
                                     &                & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
Wave$^2$                             &                & $0.017^{***}$  & $0.018^{***}$  & $0.018^{***}$  & $0.018^{***}$  & $0.018^{***}$  & $0.018^{***}$  \\
                                     &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
Perception inequality (WE)           &                &                & $-0.034^{***}$ & $-0.029^{**}$  & $-0.029^{**}$  & $-0.029^{**}$  & $-0.029^{**}$  \\

@@

Merit: Effort (WE)                   &                &                &                & $0.071^{***}$  & $0.071^{***}$  & $0.070^{***}$  & $0.070^{***}$  \\

@@

Merit: Talent (WE)                   &                &                &                & $0.028^{*}$    & $0.028^{*}$    & $0.028^{*}$    & $0.027^{*}$    \\

@@

Perception inequality (BE)           &                &                &                &                & $-0.042^{*}$   & $-0.005$       & $-0.032$       \\

@@

BIC                                  & $20780.594$    & $20757.596$    & $20761.847$    & $20704.882$    & $20714.843$    & $20633.143$    & $20727.029$    \\

@@

Var: idencuesta (Intercept)          & $0.172$        & $0.207$        & $0.204$        & $0.198$        & $0.196$        & $0.172$        & $0.160$        \\
Var: Residual                        & $0.530$        & $0.497$        & $0.497$        & $0.493$        & $0.493$        & $0.492$        & $0.493$        \\
Var: idencuesta ola\_num             & $$             & $0.007$        & $0.007$        & $0.006$        & $0.006$        & $0.006$        & $0.006$        \\
Cov: idencuesta (Intercept) ola\_num & $$             & $-0.017$       & $-0.017$       & $-0.014$       & $-0.014$       & $-0.013$       & $-0.013$       \\

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\section{Theoretical views on market justice, inequality perception, and
meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}

@@

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit and effort. It is possible to trace
the concept to the understanding of Lane
(\citeproc{ref-lane_market_1986}{1986}), who defines market justice as a
system of ``earned deserts,'' whereby individuals are seen as deserving
of a determined distributive outcome due to their effort and skills. In
contrast, political justice, emphasizes principles of equality and need,
which are often represented by the welfare state action through social
policies. An important remark is that the principles of market justice
rely on the assumption that markets are neutral and self-regulating
arenas, where individuals are treated fairly because they face the same
formal rules of engagement (\citeproc{ref-lane_market_1986}{Lane,
1986}). Consequently, the legitimacy of market justice stems from the
assumption that inequalities are not only inevitable but fair---so long
as the rules are transparent and opportunities are open. In this way,
market justice provides a moral justification for inequality by framing
it as a necessary outcome of individual responsibility.

Empirical studies have shown different strategies for the study of
market justice preferences. A common approach in the literature is to
gauge attitudes toward the legitimacy of inequality in specific domains,
especially when linked to income differences. This can be traced to the
seminal work of Kluegel and Smith
(\citeproc{ref-kluegel_beliefs_1981}{1981}) who assessed the normative
foundations that explain public support for economic inequality. Over
time, this approach has been extended beyond income to include other
market-mediated outcomes, such as education, healthcare, and/or
pensions. For example, Von Dem Knesebeck et al.
(\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
citizens consider it fair that individuals with higher incomes can
access better healthcare, while Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
context of education in Australia. These studies usually rely on a
survey item asking respondents to evaluate the fairness of income-based
access to welfare services, allowing for comparing justice perceptions
across different contexts. Similarly, comparative studies have also
relied on these indicators to study the support for market-based
distribution in welfare systems (\citeproc{ref-lindh_public_2015}{Lindh,
2015}; \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). More
recently, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. These empirical strategies all aim to capture the
extent to which individuals accept inequality when framed as a
reflection of market outcomes.

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is empirical evidence suggesting that
individuals in more advantaged socioeconomic positions---those with
higher occupational class, income, and education---are more likely to
support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
2007}). This tendency reflects not only material self-interest but also
a broader moral economy, in which winners of the market system
internalize justifications for the status quo. At the same time,
political ideology also plays a role ---such as economically
conservative values --- where more right-wing individuals show higher
support for meritocracy and more skepticism towards redistribution. This
is particularly salient in countries with more restricted public
provision of social services. For example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
those with right-leaning individuals are more supportive of market-based
distribution of welfare. Beyond individual characteristics,
country-level institutions also play a central role. In liberal welfare
regimes like those of the United States or the United Kingdom, market
justice preferences are more widespread, while in coordinated or
social-democratic regimes---such as Sweden or Germany---citizens are
generally more critical of market-based inequalities
(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
that market justice is also shaped contextually, which is represented by
the development of welfare institutions.

@@

Perceptions of inequality have been associated with attitudes about
market justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al.,
1995}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research
indicates that lower perceived inequality can reinforce support for
market-based distributive arrangements by suggesting that the system is
fair and that outcomes reflect effort and ability
(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
is perceived as excessive or structurally determined, individuals could
question the legitimacy of market justice and become more supportive of
redistributive policies
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}).

Perceptions of inequality refer to individuals' subjective evaluations
of the extent, causes, and consequences of income and wealth
disparities. Unlike objective measures such as the Gini index, perceived
inequality captures how individuals make sense of distributive
hierarchies in their everyday lives, shaped by reference groups, social
comparisons, and information environments
(\citeproc{ref-garcia-castro_perceiving_2020}{García-Castro et al.,
2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016}).
Scholars have proposed multiple dimensions of perceived inequality,
including its magnitude (how significant are the gaps), vertical
structure (between which groups), the trend over time (increasing or
decreasing), and legitimacy (whether it is just or not)
(\citeproc{ref-engelhardt_what_2018}{Engelhardt \& Wagener, 2018};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}). These dimensions encompass both cognitive and normative aspects
of perceptions of inequality and can vary across societies and social
groups, depending on exposure, ideology, and personal experience
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
2018}).

@@

Meritocracy constitutes a central ideological framework for legitimizing
different types of social inequality, for instance through market
justice beliefs. Rooted in the belief that rewards and positions should
be allocated based on individual effort and talent, meritocracy operates
as a normative ideal and a descriptive belief about how society
functions. As initially conceptualized by Michael Young
(\citeproc{ref-young_rise_1962}{1962}), the term was meant to critique a
system in which merit-based stratification becomes a new form of
inequality. However, over time, meritocracy has been widely supported in
many societies as a fair and desirable principle of distribution,
particularly within liberal democracies and market-oriented economies
(\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
perspective, the belief in meritocracy is more than a cognitive
assessment; it reflects a moral lens through which individuals interpret
inequality. People who believe that success results from hard work and
talent are more likely to view social and economic disparities as
legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
Conversely, if they see outcomes as driven by luck, social origin, or
systemic barriers, inequality is more likely to be perceived as unjust.
This distinction becomes crucial in societies with persistent structural
inequality, where public narratives often emphasize personal
responsibility and merit while overlooking entrenched disadvantages.

@@

These dimensions of meritocracy reflect how respondents perceive
society's distributive logic, regardless of whether they endorse
meritocratic principles. This distinction aligns with recent findings
indicating that individuals distinguish between how merit is perceived
in society and how it should ideally operate, which in turn shapes their
preferences for redistribution and justice
(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
2025}) Meritocratic beliefs serve as symbolic justifications for unequal
outcomes, particularly when access is stratified by income or social
opportunity. Prior studies have shown that individuals who perceive
higher levels of meritocracy tend to express stronger support for
unequal distributions that reflect market outcomes
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).

In addition to influencing individual attitudes toward inequality,
meritocratic beliefs can contribute to social division and the
stigmatization of disadvantaged groups. Recent research has demonstrated
that exposure to meritocratic narratives can reinforce the belief that
poverty results from individual failings rather than systemic
conditions, reducing support for redistributive measures and increasing
the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
et al., 2023}). This reinforces negative stereotypes and reduces empathy
toward individuals from lower socioeconomic backgrounds. Moreover,
Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
that meritocratic narratives can serve as feedback mechanisms that shape
public opinion and well-being by framing individuals' understanding of
welfare outcomes as deserved or undeserved within existing institutional
structures. This psychological mechanism highlights the normative power
of meritocracy in stabilizing unequal systems by shaping political
attitudes and personal perceptions of success and failure.

@@

\(H1\): Higher perceived economic inequality is associated with higher
market justice preferences.

\(H2\): Higher meritocratic beliefs are associated with higher market
justice preferences.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3367}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3367}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2143}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1122}}@{}}
\caption{Dependent variables for the first wave
(2016)}\label{tbl-summary1}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Label
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Stats / Values
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Freqs (\% of Valid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Valid
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Label
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Stats / Values
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Freqs (\% of Valid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Valid
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Health distributive justice &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
558 (37.2\%)\\
729 (48.6\%)\\
63 ( 4.2\%)\\
133 ( 8.9\%)\\
18 ( 1.2\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
Pension distributive justice &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
426 (28.4\%)\\
718 (47.8\%)\\
108 ( 7.2\%)\\
226 (15.1\%)\\
23 ( 1.5\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
Education distributive justice &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
521 (34.7\%)\\
783 (52.2\%)\\
73 ( 4.9\%)\\
113 ( 7.5\%)\\
11 ( 0.7\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
Market justice preferences & \begin{minipage}[t]{\linewidth}\raggedright
Mean (sd) : 2 (0.8)\\
min \textless{} med \textless{} max:\\
1 \textless{} 2 \textless{} 5\\
IQR (CV) : 0.7 (0.4)\strut
\end{minipage} & 12 distinct values &
\begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
\end{longtable}

@@

Intercept                            & $1.968^{***}$  & $2.015^{***}$  & $2.030^{***}$  & $2.028^{***}$  & $2.169^{***}$  & $1.456^{***}$  & $1.547^{***}$  \\
                                     & $(0.021)$      & $(0.035)$      & $(0.035)$      & $(0.035)$      & $(0.070)$      & $(0.097)$      & $(0.115)$      \\

@@

Wave                                 &                & $-0.121^{***}$ & $-0.125^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.129^{***}$ \\
                                     &                & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      \\
Wave$^2$                             &                & $0.028^{***}$  & $0.028^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  \\
                                     &                & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
Perception inequality (WE)           &                &                & $-0.039^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ \\

@@

Merit: Effort (WE)                   &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.069^{***}$  & $0.069^{***}$  \\

@@

Merit: Talent (WE)                   &                &                &                & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    \\

@@

Perception inequality (BE)           &                &                &                &                & $-0.042^{*}$   & $-0.006$       & $-0.033$       \\

@@

BIC                                  & $20780.594$    & $20780.617$    & $20781.113$    & $20726.753$    & $20736.627$    & $20655.172$    & $20749.047$    \\

@@

Var: idencuesta (Intercept)          & $0.172$        & $0.223$        & $0.219$        & $0.209$        & $0.207$        & $0.182$        & $0.171$        \\
Var: Residual                        & $0.530$        & $0.499$        & $0.498$        & $0.495$        & $0.495$        & $0.495$        & $0.495$        \\
Var: idencuesta ola\_num             & $$             & $0.011$        & $0.010$        & $0.009$        & $0.009$        & $0.009$        & $0.009$        \\
Cov: idencuesta (Intercept) ola\_num & $$             & $-0.025$       & $-0.024$       & $-0.021$       & $-0.020$       & $-0.019$       & $-0.019$       \\
